Imputation of Incomplete Multilevel Data with R

### 2023-12-11 docs/manuscript.tex
Add complete case study 1 and 2, and text of 3

\begin{verbatim}
Warning: package 'ggplot2' was built under R version 4.3.2
\end{verbatim}

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomrule=.15mm, toprule=.15mm, breakable, leftrule=.75mm, left=2mm, colback=white, opacityback=0, rightrule=.15mm]

@@

\begin{verbatim}
plot_na()
\end{verbatim}

\begin{figure}[t]

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomrule=.15mm, toprule=.15mm, breakable, leftrule=.75mm, left=2mm, colback=white, opacityback=0, rightrule=.15mm]

Box 3. The predictor matrix. The entries corresponding to the level-1
predictors are coded with a 3, indicating that both the original values
as well as the cluster means of the predictor are included into the
imputation model. The entry of 4 in the predictor matrix adds three
variables to the imputation model for the imputation model predictor:
the value of the predictor, the cluster means of the predictor and the
random slopes of the predictor.

@@

@@

\begin{verbatim}
R> set.seed(123)
R> library(mice)
R> library(ggmice)
R> library(ggplot2)
R> library(miceadds)
R> library(lme4)
R> library(mitml)
R> library(broom.mixed)
\end{verbatim}

@@

@@

\begin{verbatim}
R> dat <- popmis[, c("school", "teachpop", "popular", "texp", "sex")] 
\end{verbatim}

\begin{verbatim}
ggmice(dat, aes(popular, teachpop)) + 
  geom_jitter()
\end{verbatim}

\begin{figure}[t]

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-7-1.pdf}

@@

\caption{Polar axis plot}

@@

With the \texttt{ggmice} unction \texttt{plot\_pattern} we can visualize
this.

@@

\begin{figure}[t]

@@

\begin{figure}

@@

@@

\begin{verbatim}
  school teachpop  popular     texp      sex 
      ""       ""    "pmm"       ""       "" 
\end{verbatim}

@@

\begin{verbatim}
         school teachpop popular texp sex
school        0        0       0    0   0
teachpop      0        0       0    0   0
popular       0        1       0    1   1
texp          0        0       0    0   0
sex           0        0       0    0   0
\end{verbatim}

@@

@@

\begin{figure}[t]

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-15-1.pdf}

@@

\begin{figure}[t]

@@

\begin{verbatim}
ggmice(imp, aes(popular, group = .imp)) + 
  geom_density() 
\end{verbatim}

\begin{figure}[t]

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-18-1.pdf}

@@

\begin{verbatim}
ggmice(imp, aes(.imp, popular)) + 
  geom_jitter(alpha = 0.05) +
    geom_boxplot()
\end{verbatim}

\begin{figure}[t]

\begin{verbatim}
ggmice(imp, aes(popular, teachpop)) + 
  geom_jitter() +
  facet_wrap(~ .imp)
\end{verbatim}

\begin{figure}[t]

@@

\begin{verbatim}
fit <- with(
  imp,
  lmer(teachpop ~ popular + texp + (1 | school))
)
\end{verbatim}

Pool the estimates.

\begin{verbatim}
pool(fit)
\end{verbatim}

\begin{verbatim}
Class: mipo    m = 5 
         term m  estimate         ubar            b            t dfcom
1 (Intercept) 5 2.4091354 2.241304e-02 1.712964e-03 2.446860e-02  1995
2     popular 5 0.2597284 2.353344e-04 1.209648e-04 3.804922e-04  1995
3        texp 5 0.0484727 7.728295e-05 3.236252e-06 8.116646e-05  1995
         df        riv     lambda        fmi
1 432.50579 0.09171257 0.08400798 0.08821454
2  26.88403 0.61681474 0.38149995 0.42289329
3 909.68447 0.05025044 0.04784615 0.04993264
\end{verbatim}

@@

\begin{verbatim}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{verbatim}

\begin{verbatim}

Call:

testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)

Final parameter estimates and inferences obtained from 5 imputed data sets.

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     2.409     0.156    15.401   566.786     0.000     0.092     0.087 
popular         0.260     0.020    13.315    27.483     0.000     0.617     0.422 
texp            0.048     0.009     5.380  1747.294     0.000     0.050     0.049

                            Estimate 
Intercept~~Intercept|school    0.310 
Residual~~Residual             0.307 
ICC|school                     0.502

Unadjusted hypothesis test as appropriate in larger samples.
\end{verbatim}

>>>>

In hierarchical datasets, clustering is a concern because the
homoscedasticity in the error terms cannot be assumed across clusters
and the relationship among variables may vary at different hierarchical
levels. When multiple imputation is used to deal with missing data, as
the imputation and analysis process is performed separately, it is
necessary that imputation model being congenial with the main analysis
model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
structure also imputation model should do it (Audigier, 2021). Not
including clustering into the imputation process may lead to effect
estimates with smaller standard errors and inflated type I error.

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, arc=.35mm, colback=white, breakable, toprule=.15mm, bottomrule=.15mm, rightrule=.15mm, leftrule=.75mm, left=2mm]

@@

\begin{figure}[h]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, arc=.35mm, colback=white, breakable, toprule=.15mm, bottomrule=.15mm, rightrule=.15mm, leftrule=.75mm, left=2mm]

Box 3. The predictor matrix. The entries corresponding to the level-1
predictors are coded with a 3, indicating that both the original values
as well as the cluster means of the predictor are included into the
imputation model. The entry of 4 in the predictor matrix adds three
variables to the imputation model for the imputation model predictor:
the value of the predictor, the cluster means of the predictor and the
random slopes of the predictor. - -2 = cluster variable - 1 = overall
effect - 3 = overall + group-level effect - 4 = individual-level
(random) and group-level (fixed) effect

@@

\hypertarget{sec-workflow}{%
\section{Multilevel imputation workflow}\label{sec-workflow}}

There are different strategies that can be adopted in the imputation
process that account for clustering: inclusion of cluster indicator
variable, performing a separate imputation process for each cluster, or
performing a simultaneous imputation process by using an imputation
method that accounts for clustering.(Stata:
https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/)
TODO: replace ref.

The selection of each strategy depends mainly on the assumptions in the
main analysis and also on the restriction of the analyzed data.

Regarding the restrictions imposed by the data, for instance, the use of
cluster indicator variables is restricted in datasets where there are
not many clusters and many observations per cluster (Graham, 2009). The
last restriction is also required when imputations are performed on each
cluster separately. When this restriction cannot be achieved, one can
use an imputation model that simultaneously imputes all clusters using a
hierarchical model (Allison 2002).

Under this hierarchical imputation model, observations within clusters
are correlated and this correlation is modeled by a random effect so the
hierarchical model can be estimated even when there are few observations
per cluster. However, this strategy is best suited for balanced data
(Grund, 2017) and when random effects model is appropriated, i.e.~the
number of clusters is adequate. (Austin,2018).

Here it is important to evaluate the assumptions imposed by the main
model, for instance by using the cluster indicator strategy may lead to
bias estimates when the model is based on a hierarchical model
(Taaljard,2008). Even when an imputation strategy congenial with the
main model is preferred, it is important to consider whether it is
appropriate for the data as a less complex imputation strategies may
also lead to unbiased estimates in certain scenarios(Bailey 2020). For
instance, in causal effect analysis, separately imputation may lead to
smaller bias when the size of the smaller exposure cluster is large,
compared with an imputation model that includes exposure-confounder
interactions. (Zhang,2023).

Below we provide a imputation workflow that can be used in general to
impute cluster data.

\hypertarget{focus-on-the-main-analysis}{%
\subsection{Focus on the Main
Analysis}\label{focus-on-the-main-analysis}}

When imputing clustered data, we initially focus on the research
question and the intended analysis, assuming the presence of
non-incomplete values. This assessment sheds light on research
hypotheses and the contemplated main(s) statistical model(s) and it
provides insights into the data's structure (refer to the level table),
variable types (e.g., confounders, auxiliary variables), and other
considerations about variables relationships such as interactions or
polynomial terms.

\hypertarget{exploration-of-available-data}{%
\subsection{Exploration of Available
Data}\label{exploration-of-available-data}}

Next, it is necessary to explore the information available in the data.
Exploring individual variables using tools such as histograms or QQ
plots can help to delineate variable distributions and plausible ranges
of values and also identify input errors or outliers. Evaluating
interactions between predictors using scatter plots or conditional
plots, assessing collinearity using VIF or correlations, can help to
glimpse nonlinear relationships that may affect the original formulation
of the main model. Further testing of assumptions related to the
response variable, including variance (homogeneity using conditional box
plots), independence (e.g., ACF, variograms) and response vs.~predictor
relationships (conditional plots - x vs.~y), may serve to choose an
imputation model more suitable to the data. In addition, the intraclass
correlation coefficient (ICC) can be examined to assess cluster
differences, aiding in the choice between the 2l and 1l methods for
imputation.

One can also explore missingness, specifically missing patterns at the
cluster level to identify types of patterns (e.g., non-monotonic,
systematic) to guide the selection of an appropriate imputation approach
in terms of computational efficiency (e.g., simpler regression
imputation versus FCS in univariate patterns). Similarly, these missing
patterns can also be used to identify potential predictors in individual
imputation models by means of inflow criteria (quantifying the
connections between missing data in one variable and other observed
variables) and outflow criteria (determining the connections between
observed values in one variable and missing data in other variables).
Although several packages exist for visualizing missing data, this
tutorial focuses on the recently implemented ggmice package.

\hypertarget{assess-estimation-procedure-robustness-to-missing-data}{%
\subsection{Assess Estimation Procedure Robustness to Missing
Data}\label{assess-estimation-procedure-robustness-to-missing-data}}

Prior to implementing multiple imputation, it is critical to assess the
robustness of the estimation model when confronted with missing data.
There are scenarios in which specific Maximum Likelihood (ML) estimation
methods outperform Multiple Imputation (MI) methods. For instance, when
the response variable is the sole incomplete variable, mixed models
demonstrate robustness to missing data under the Missing at Random (MAR)
assumption and with a correct variance-covariance
specification.\cite{Molenberghs_2007}. Furthermore, depending on the
proportion of missingness (usually is \textless5\%), opting for a
simpler complete case analysis might be sufficient.

\hypertarget{pre-imputation}{%
\subsection{Pre-imputation}\label{pre-imputation}}

Before proceeding with the imputation model, it is necessary to filter
the dataset to include only the minimal variables required for the main
model. If additional procedures are conducted during the analysis, it is
essential to include variables associated with these procedures, such as
confounders in the case of balancing techniques. Other relevant
variables, like instrumental variables or auxiliary variables, may
enhance parameter estimates even if not included in the main model, as
they could be linked to the probability of missingness for some
incomplete variables.

Furthermore, at the variable level, it is crucial to assess whether
proxy variables should be considered or if deterministic imputation
alone is adequate for imputing missing values. For instance, certain
incomplete variables may not necessitate stochastic imputation methods
like MICE. Instead, they can be effectively addressed through deductive
imputation, where incomplete values are inferred from logical and
deterministic relationships between variables. This approach is
especially beneficial for variables that are functions of others, such
as deriving a person's BMI from their weight and height. It also proves
valuable for determining values for one-level variables from two-level
ones, as seen in the context of Individual Participant Data (IPD). In
such cases, incomplete information can be deduced from metadata, like
inferring incomplete data about abortion in a country where abortion is
illegal, or through cross-temporal or protocol-based deduction, such as
imputing missing test values for deceased patients.

\hypertarget{setting-imputation-model}{%
\subsection{Setting Imputation Model}\label{setting-imputation-model}}

\hypertarget{clustering-inclusion}{%
\paragraph{Clustering Inclusion}\label{clustering-inclusion}}

Various strategies can be employed in the imputation process to account
for clustering, including the inclusion of a cluster indicator variable,
conducting a separate imputation process for each cluster, or employing
a simultaneous imputation method that considers clustering\cite{eddings}

The choice of strategy depends primarily on the assumptions made in the
main analysis and the constraints imposed by the analyzed data.
Concerning data restrictions when the analysis do not contemplate an
hierarchical model (eg. descriptive way), if there are few clusters with
many observations per cluster, the cluster indicator or imputation
separated by groups may be appropriate \cite{graham2009}. Conversely, an
imputation model that simultaneously imputes all clusters using a
hierarchical model can be employed when there are more clusters or fewer
observations per cluster \cite{Allison_2002}.

In a hierarchical imputation model, correlations between observations
within clusters are modeled by a random effect, allowing estimation even
with a limited number of observations per cluster. Multiple imputation
models based on hierarchical models have been proposed, each with
different assumptions. \cite{audigier}

In the particular case that the analysis contemplates the use of a
hierarchical model (i.e.~main model), before choosing a particular
strategy, it is crucial to verify if the assumptions of the imputation
model align with those of the main analysis. For example, using the
cluster indicator strategy may introduce bias estimates when the model
is based on a hierarchical structure \cite{taljaard2008,speidel2018}.
Even if an imputation strategy congruent with the main model is
preferred, it is essential to assess its appropriateness for the data,
considering that less complex imputation strategies may lead to unbiased
estimates in certain scenarios \cite{bailey2020}.

\hypertarget{choice-of-individual-imputation-methods}{%
\subsubsection{Choice of Individual Imputation
Methods}\label{choice-of-individual-imputation-methods}}

The initial step involves selecting the individual imputation model for
each incomplete variable in the dataset. The mice package automatically
proposes imputation methods based on the variable type for 1l variables.
For 2l level variables, the micemd package's () function can be
utilized. This function selects among the 2l imputation methods based on
the size of the clusters and the proportion of missingness in each
cluster.

In addition to the package-defined imputation methods, users can specify
custom methods using the ``I formula''. This flexibility allows for the
calculation of derived variables internally or adjustments to imputation
methods based on specific conditions, such as conditioning the
imputation model on the level of an incomplete covariate (e.g.,
pregnancy test on females).

\hypertarget{model-specification}{%
\subsubsection{Model Specification}\label{model-specification}}

The imputation model must be congenial with the main model
\cite{meng1994}. Congeniality issues arise when the imputation model and
the main model make different assumptions, often due to the omission of
a polynomial or interaction term or the use of transformed variables.

The imputation model can include additional terms than the main model
terms that do not lead to uncongeniality. For instance, it is advisable
to include the outcome variable in the imputation model for prediction
variables \cite{moons2006a}. In cases where the outcome is
time-to-event, the Nelson-Aalen estimate of the time to the event should
be included as a covariate in the imputation model {[}REF{]}.
Furthermore, the inclusion of auxiliary variables, even if not part of
the main model, can be associated with the probability of missingness,
enhancing the likelihood of satisfying the Missing at Random (MAR)
assumption and improving estimation efficiency \cite{hardt2012a}.

The specification of imputation models is done on a variable basis,
either using a prediction matrix where the type of prediction variable
should be specified accordingly (see table) or through a list of
formulas in the formula parameter. The latter option proves useful in
formulating imputation models with polynomial terms or interactions
compared to the prediction matrix specification, which requires the
inclusion of additional terms as Just Another Variable (JAV)
\cite{buuren2018a}.

However, even when considering an interaction term, depending on the
adapted individual imputation model, it has been suggested, for
instance, for treatment interaction effects, to conduct separated
imputation by treatment group \cite{zhang2023}. Additionally, there are
imputation models based on random forest or deep learning that can
handle interaction and non-linear terms that do not require the explicit
inclusion of the non-linear terms.

\hypertarget{post-imputation}{%
\subsection{Post-Imputation}\label{post-imputation}}

During the imputation process, certain issues may arise that halt the
process. In hierarchical model imputations, many issues are linked to
methods based on parametric hierarchical models, which may struggle to
estimate a model midway through the process. In such cases, it is
advisable to examine the imputation log file to identify variables with
imputation issues. Subsequently, reducing the number of predictors in
the imputation model could be explored, either by using functions like
quickpred or by considering variable transformations, such as scaling.
Adjusting the level of the hierarchical model (e.g., using a homogenous
variance model or a 1l model) can also be beneficial. Additionally,
checking the range of imputed variables is crucial, as excessively large
imputed values for one predictor could lead to convergence issues in
other variables. This can be mitigated by including post-processing
specifications on problematic variables or by using imputation models
like Predicted Mean Matching (PMM), ensuring imputed values align with
the observable values. In certain cases, a separate imputation strategy
may be considered. For instance, in analyses involving multiple
endpoints, conducting separate imputation processes for each endpoint
might be preferable to a unified imputation process.

\hypertarget{convergence-and-sensitivity-analysis}{%
\subsection{Convergence and Sensitivity
Analysis}\label{convergence-and-sensitivity-analysis}}

Before starting into the analysis of each imputed dataset, it is crucial
to validate the convergence of the imputation process. This is commonly
accomplished through trace plots that depict the mean and variance of
the incomplete variables across iterations. These plots serve to uncover
potential circular issues or the need for additional iterations.
Additionally, it is also important to verify that imputed values fall
within a plausible range and also to check the distribution of imputed
variables, ensuring that the imputed variable distribution aligns with
the distribution of observed values (under the MAR assumption). An
alternative approach involves assessing the prediction accuracy of the
imputation method \cite{cai2023}.

While the majority of Multiple Imputation by Chained Equations (MICE)
methods are based on Missing at Random (MAR) assumptions, field expert
input may suggest that the Missing Not at Random (MNAR) mechanism could
be plausible for certain variables. An MNAR variable is one in which the
probability of missingness depends on an unobservable variable. This can
occur when missingness is associated with the incomplete value itself
(self-marking) or when there is an unobserved variable linked to both
the value and the probability of missingness of the incomplete value
(indirectly non-informative). Specifically, for the indirectly
non-informative case in hierarchical datasets, imputation methods based
on the Heckman method can be
considered.\cite{hammon2020,hammon2022,munoz2023}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

@@

\begin{verbatim}
R> set.seed(123)         # for reproducibility
R> library(lme4)         # for multilevel modeling
R> library(mice)         # for imputation
R> library(miceadds)     # for multilevel imputation methods
R> library(micemd)       # for selection-model imputation methods
R> library(mitml)        # for multilevel parameter pooling
R> library(broom)        # for clean model estimates
R> library(broom.mixed)  # for multilevel model estimates
R> library(dplyr)        # for data wrangling
R> library(ggmice)       # for visualization
R> library(ggplot2)      # for visualization
\end{verbatim}

@@

In this section we will go over the different steps involved with
imputing incomplete multilevel data with the R package mice. We consider
the simulated \texttt{popmis} dataset, which included pupils
(\(n = 2000\)) clustered within schools (\(N = 100\)). The following
variables are of primary interest:

\begin{itemize}
\tightlist
\item
  \texttt{school}, school identification number (clustering variable);
\item
  \texttt{popular}, pupil popularity (self-rating between 0 and 10;
  unit-level);
\item
  \texttt{sex}, pupil sex (0 = boy, 1 = girl; unit-level);
\item
  \texttt{texp}, teacher experience (in years; cluster-level).
\end{itemize}

The analysis model corresponding to this dataset is multilevel
regression with random intercepts for the different schools. We will
estimate the association between the pupils' sex and their popularity
score. This model can be expressed in \texttt{lme4} code as:

\begin{verbatim}
popular ~ 1 + sex + (1 | school)
\end{verbatim}

We load the data into the environment with

@@

and select the relevant variables

\begin{verbatim}
R> dat <- popmis[, c("school", "popular", "texp", "sex")] 
\end{verbatim}

which results in the following data structure.

\begin{verbatim}
R> head(dat)
\end{verbatim}

\begin{verbatim}
  school popular texp sex
1      1      NA   24   1
2      1      NA   24   0
3      1       7   24   1
4      1      NA   24   1
5      1      NA   24   1
6      1       7   24   0
\end{verbatim}

The association of interest can be visualized with \texttt{ggmice},

\begin{verbatim}
ggmice(dat, aes(sex, popular)) +
  geom_jitter()
\end{verbatim}

\begin{figure}[h]

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-8-1.pdf}

@@

\caption{Scatterplot of student popularity by sex}

@@

where missing datapoints in the \texttt{popular} variable are
represented by red points on the X-axis of the figure.

With the \texttt{ggmice} function \texttt{plot\_pattern} we can
visualize the missing data pattern

@@

\begin{figure}[h]

@@

which shows us that the missingness is univariate and sporadic.

To develop the best imputation model for the incomplete variable
\texttt{popular}, we need to know whether the observed values of
\texttt{popular} are related to observed values of other variables. Plot
the pair-wise complete correlations in the incomplete data

@@

This shows us that not just the analysis-model variable \texttt{sex},
but also the cluster-level covariate teacher experience, \texttt{texp},
may be a useful as an imputation model predictor. Moreover, the
missingness in \texttt{popular} may depend on the observed values of
other variables. With \texttt{ggmice()} we can visualize the
distribution of the teacher experience for cases where \texttt{popular}
is observed and cases where \texttt{popular} is missing.

\begin{verbatim}
R> ggmice(dat, aes(texp)) + 
+  geom_density() +
+  facet_wrap(~is.na(popular))
\end{verbatim}

\begin{figure}[h]

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-11-1.pdf}

It appears that students with a missing value for \texttt{popular} are
in clusters with a slightly higher \texttt{texp} value.

\begin{verbatim}
t.test(dat$texp ~ is.na(dat$popular)) |> 
  tidy() |>
  kable()
\end{verbatim}

\begin{longtable}[]{@{}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0924}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0924}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0924}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.0840}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.2017}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 18\tabcolsep) * \real{0.1008}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedleft
estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
estimate1
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
estimate2
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
statistic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
p.value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
parameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
conf.low
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
conf.high
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
method
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
alternative
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
-0.2599581 & 14.15278 & 14.41274 & -0.8731291 & 0.3827095 & 1796.115 &
-0.8438947 & 0.3239786 & Welch Two Sample t-test & two.sided \\
\end{longtable}

Although there are no significant differences in the distribution of
\texttt{texp} depending on the missingness indicator of
\texttt{popular}, this variable can serve as auxiliary variable in the
imputation of \texttt{popular}.

@@

\begin{verbatim}
 school popular    texp     sex 
     ""   "pmm"      ""      "" 
\end{verbatim}

@@

\begin{verbatim}
        school popular texp sex
school       0       0    0   0
popular      0       0    1   1
texp         0       0    0   0
sex          0       0    0   0
\end{verbatim}

@@

The \texttt{pmm} method is better (more efficient) because it will still
look for donors (maybe outside of cluster) based on predictive distance,
even for very small clusters.

@@

\begin{figure}[h]

@@

\begin{figure}[h]

Troubleshoot non-convergence of imputation model. Evaluate the
distribution of imputed values.

\begin{verbatim}
R> ggmice(imp, aes(popular, group = .imp)) + 
+  geom_density() 
\end{verbatim}

\begin{figure}[h]

@@

\begin{verbatim}
R> ggmice(imp, aes(.imp, popular)) + 
+  geom_jitter(alpha = 0.05) +
+    geom_boxplot()
\end{verbatim}

\begin{figure}[h]

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-21-1.pdf}

@@

\begin{verbatim}
R> ggmice(imp, aes(sex, popular)) +
+  geom_jitter() +
+  facet_wrap(~ .imp)
\end{verbatim}

\begin{figure}[h]

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-22-1.pdf}

@@

\begin{verbatim}
R> fit <- with(
+  imp,
+  lmer(popular ~ texp + sex + (1 | school))
+)
\end{verbatim}

Pooling the estimates does not provide estimates of the variance
components.

\begin{verbatim}
R> pool(fit)
\end{verbatim}

\begin{verbatim}
Class: mipo    m = 5 
         term m   estimate         ubar            b            t dfcom
1 (Intercept) 5 3.59902513 0.0284760238 3.461284e-03 0.0326295651  1995
2        texp 5 0.09225928 0.0001140591 2.273382e-05 0.0001413397  1995
3         sex 5 0.85187018 0.0009628869 4.946825e-04 0.0015565060  1995
        df       riv    lambda       fmi
1 216.1756 0.1458610 0.1272938 0.1352573
2 100.6502 0.2391794 0.1930144 0.2085857
3  26.9008 0.6164992 0.3813792 0.4227574
\end{verbatim}

Therefore, \texttt{mitml} is used.

\begin{verbatim}
R> est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{verbatim}

@@

\begin{verbatim}
R> est$estimates |> 
+  round(3) |>
+  kable()
\end{verbatim}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1558}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1169}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1299}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2338}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Std.Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
t.value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
df
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
P(\textgreater\textbar t\textbar)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
RIV
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
FMI
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
(Intercept) & 3.599 & 0.181 & 19.924 & 246.857 & 0 & 0.146 & 0.134 \\
texp & 0.092 & 0.012 & 7.760 & 107.369 & 0 & 0.239 & 0.208 \\
sex & 0.852 & 0.039 & 21.592 & 27.501 & 0 & 0.616 & 0.422 \\
\end{longtable}

\begin{verbatim}
R> est$extra.pars |> 
+  round(3) |>
+  kable()
\end{verbatim}

\begin{longtable}[]{@{}lr@{}}
\toprule\noalign{}
& Estimate \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Intercept\textasciitilde\textasciitilde Intercept\textbar school &
0.470 \\
Residual\textasciitilde\textasciitilde Residual & 0.462 \\
ICC\textbar school & 0.504 \\
\end{longtable}

\hypertarget{impact-data}{%
\subsection{IMPACT data}\label{impact-data}}

The second case study is the \texttt{impact} data from the
\pkg{metamisc} package \citep[empirical data on traumatic brain
injuries, \(n = 11,022\) units across \(N = 15\) clusters,][]{metamisc}.

The \texttt{impact} data set contains traumatic brain injury data on
\(n = 11022\) patients clustered in \(N = 15\) studies with the
following 11 variables:

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification,
\item
  \texttt{hypox} Hypoxia (0=no, 1=yes),
\item
  \texttt{hypots} Hypotension (0=no, 1=yes),
\item
  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
\item
  \texttt{edh} Epidural hematoma (0=no, 1=yes),
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

Check if there is systematic missingness in this dataset. For
illustration purposes, we made Marshall Computerized Tomography
classification (ct) systematically missing.

The analysis model for this dataset is a prediction model with
\texttt{mort} as the outcome. In this tutorial we'll estimate the
adjusted prognostic effect of \texttt{ct} on mortality outcomes. The
estimand is the adjusted odds ratio for \texttt{ct}, after including
\texttt{type}, \texttt{age} \texttt{motor\_score} and \texttt{pupil}
into the analysis model:

\begin{verbatim}
mort ~ type + age + motor_score + pupil + ct + (1 | name) 
\end{verbatim}

Note that variables \texttt{hypots}, \texttt{hypox}, \texttt{tsah} and
\texttt{edh} are not part of the analysis model, and may thus serve as
auxiliary variables for imputation.

The \texttt{impact} data included in the \pkg{metamisc} package is a
complete data set. The original data has already been imputed once
(Steyerberg et al, 2008). For the purpose of this tutorial we have
induced missingness (mimicking the missing data in the original data set
before imputation). The resulting incomplete data can be accessed from
\href{https://zenodo.com}{zenodo link to be created}.

Load the incomplete data into the R workspace:

\begin{verbatim}
R> dat <- read.table("link/to/the/data.txt") 
\end{verbatim}

To explore the missingness, we should look at the missing data pattern.
The ten most frequent missingness patterns are shown with

\begin{verbatim}
R> plot_pattern(dat, rotate = TRUE, npat = 10L)  
\end{verbatim}

\begin{figure}[h]

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-28-1.pdf}

This shows that we need to impute \texttt{ct} and \texttt{pupil}. To
develop the best imputation model, we need to investigate the relations
between the observed values of the incomplete variables and the observed
values of other variables, and the relation between the missingness
indicators of the incomplete variables and the observed values of the
other variables. To see whether the missingness depends on the observed
values of other variables, we can test this statistically or use visual
inspection (e.g.~a histogram faceted by the missingness indicator).

We should impute the variables \texttt{ct} and \texttt{pupil} and any
auxiliary variables we might want to use to impute these incomplete
analysis model variables. We can evaluate which variables may be useful
auxiliaries by plotting the pairwise complete correlations

\begin{verbatim}
R> plot_corr(dat, rotate = TRUE)
\end{verbatim}

\begin{figure}[h]

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-29-1.pdf}

This shows us that \texttt{hypox} and \texttt{hypot} would not be useful
auxiliary variables for imputing \texttt{ct}. Depending on the minimum
required correlation, \texttt{tsah} could be useful, while \texttt{edh}
has the strongest correlation with \texttt{ct} out of all the variables
in the data and should definitely be included in the imputation model.
For the imputation of \texttt{pupil}, none of the potential auxiliary
variables has a very strong relation, but \texttt{hypots} could be used.
We conclude that we can exclude \texttt{hypox} from the data, since this
is neither an analysis model variable nor an auxiliary variable for
imputation

\begin{verbatim}
R> dat <- select(dat, !hypox)
\end{verbatim}

Mutate data to get the right data types for imputation (e.g.~integer for
clustering variable).

\begin{verbatim}
R> dat <- dat |>
+  mutate(name = as.integer(name))
\end{verbatim}

Create an initial methods vector for the incomplete variables

\begin{verbatim}
       name        type         age motor_score       pupil          ct 
         ""          ""          ""          ""   "polyreg"   "polyreg" 
     hypots        tsah         edh        mort 
      "pmm"       "pmm"       "pmm"          "" 
\end{verbatim}

which should be adjusted to the appropriate \texttt{2l} methods.

\begin{verbatim}
R> meth[c("pupil", "ct")] <- "2l.pmm"
R> meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
\end{verbatim}

Create an initial predictor matrix

\begin{verbatim}
R> pred <- quickpred(dat)
\end{verbatim}

This predictor matrix is too large to display inline. A visualization of
the adapted predictor matrix is presented in Figure XYZ.

We should make sure \texttt{name} is used as clustering variable

\begin{verbatim}
R> pred[, "name"] <- -2
\end{verbatim}

and the analysis-model outcome should be used as a predictor in all
imputation models

\begin{verbatim}
R> pred[, "mort"] <- 2
\end{verbatim}

the resulting predictor matrix is visualized with

\begin{verbatim}
R> plot_pred(pred, method = meth, rotate = TRUE)
\end{verbatim}

\begin{figure}[h]

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-37-1.pdf}

Impute the incomplete data with

\begin{verbatim}
R> # imp <- mice(
R> #   dat,
R> #   method = meth,
R> #   predictorMatrix = pred,
R> #   printFlag = FALSE
R> # )
\end{verbatim}

Evaluate the convergence of the algorithm

\begin{verbatim}
R> # plot_trace(imp)
\end{verbatim}

Analyze the imputed data:
